---
title: "HW4 yq2378"
author: "Yumeng Qi"
date: "2024-02-27"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# 1 summarize th data 

```{r}

df1 = tibble(
  satisfaction = rep(c("low", "medium", "high"), each = 6),
  contact = rep(rep(c("low", "high"),each = 3),3),
  housing = rep(c("Tower Block", "Apartment","House"),6),
  cnt = c(65,130,67,34,141,130,54,76,48,47,116,105,100,111,62,100,191,104)) |>
  mutate(satisfaction = factor(satisfaction, levels = c("low", "medium", "high"), ordered = TRUE),
         contact = factor(contact, levels = c("high","low")) )

# Create a table for satisfaction vs contact
contact_table <- xtabs(cnt ~ satisfaction + contact, data = df1)

# Convert counts to percentages
round(prop.table(contact_table,2) * 100,2)
round(prop.table(contact_table,1) * 100,2)
# Create a table for satisfaction vs housing
housing_table <- xtabs(cnt ~ satisfaction + housing, data = df1)

# Convert counts to percentages
round(prop.table(housing_table,2) * 100,2)

```

## 1) contact with other residents and levels of satisfaction

High contact with other residents has a larger proportion of high satisfaction of residents (40.81%) compared to low contact with other residents (38.29%). However, medium satisfaction residents have the least proportion in both high and low contact with other.

## 2) type of housing and levels of satisfaction 

Those living in the block has 50% of high satifaction and nearly equal amount of low and medium satisfaction. While for those living in the house, the majority (38.18%) has low satisfaction. 




# 2


```{r summary}

df2 = tibble(
  contact = rep(c("low", "high"),each = 3),
  housing = rep(c("Tower Block", "Apartment","House"),2),
  low_satisfaction = c(65,130,67,34,141,130),
  medium_satisfaction = c(54,76,48,47,116,105),
  high_satisfaction = c(100,111,62,100,191,104)) 

library(nnet)
mult <- multinom(cbind(low_satisfaction, medium_satisfaction, high_satisfaction)~contact+housing, data=df2)
summary_mult <- summary(mult)

summary_mult
```

*For "medium satisfaction" relative to the baseline "low satisfaction"*

Being low in contact with others has a negative coefficient (-0.2960), suggesting that lower contact is associated with a decrease in the log-odds of medium satisfaction. As for housing in House and Tower Block, both have positive coefficient (0.0697 and 0.4068 respectively), suggesting an increase in the log-odds of medium satisfaction and clearly Tower Block has a stronger impact than House.


*For "high satisfaction" relative to the baseline "low satisfaction"*

Being low in contact with others has a negative coefficient (-0.3282), which means larger impact, but still indicates consistency in the direction with medium satisfaction. Housing in House has a negative coefficient (-0.3040), suggesting living in a house is associated with a decrease in the log-odds of high satisfaction.Housing in Tower Block has a positive coefficient, significantly increasing the log-odds of high satisfaction for tower block residents

```{r CI}
coefficients <- summary_mult$coefficients
std_errors <- summary_mult$standard.errors

# Calculate 95% Confidence Intervals
conf_int_lower <- coefficients - 1.96 * std_errors
conf_int_upper <- coefficients + 1.96 * std_errors

CI = paste("(",round(conf_int_lower,2),", ", round(conf_int_upper,2), ")",sep = "")

CI = rbind(CI[seq(1,8,by =2)],CI[seq(2,8,by =2)])
colnames(CI) = colnames(conf_int_lower) 
rownames(CI) = rownames(conf_int_lower)
print("95% Confidence Interval")
CI
```

From the CI, we find only housing being House and medium satisfaction has a CI contains 0. It indicates except for this combination, all other coefficients are significant under 0.05 significance value.

```{r goodness of fit}
# goodness of fit
pihat1=predict(mult,type='probs') 
m = rowSums(df2[,3:5])
res.pearson1=(df2[,3:5]-pihat1*m)/sqrt(pihat1*m) # pearson residuals
G.stat1=sum(res.pearson1^2) # Generalized Pearson Chisq Stat
G.stat1 # Generalized Pearson chisq statistic
D.stat1=sum(2*df2[,3:5]*log(df2[,3:5]/(m*pihat1)))
D.stat1 # Deviance statistics
```

As for teh goodness of fit, the Generalized Pearson chisq statistic is `r G.stat1` and the Deviance statistics is `r D.stat1`.


```{r interaction}

mult2 <- multinom(cbind(low_satisfaction, medium_satisfaction, high_satisfaction)~contact*housing, data=df2)
#summary(mult2)
anova(mult, mult2, test="Chisq")
```
Given interaction leads to a nested model, we test if the null model stands, which suggests that there is no interaction term.From the result we know the p-value is 0.1417. Under significance value of 0.05, we fail to reject the null hypothesis, believing there is no interaction of contact level by house type. 


# 3

```{r ordinal category}
library(MASS)
# fit proportional odds model
polr= polr(satisfaction~contact+housing,data=df1,weights=cnt)
summary(polr)  
```

* contact low: This coefficient (0.2524) represents the effect of the "contact" predictor on the log odds of being in a higher satisfaction category, holding other variables constant. The positive sign indicates that being contact low is associated with lower odds of being in a higher satisfaction category compared to being contact high. This trend is consistent with nominal regression. The t-value of -2.713 and the corresponding standard error suggest that this effect is statistically significant.

* housing House: The coefficient (0.2353) for "housing House" indicates that living in a house, compared to the baseline housing category, Apartment and among those with same level of contact, is associated with lower odds of being in a higher satisfaction category. The negative sign denotes a decrease in the odds of higher satisfaction for residents in houses, and the t-value of -2.236 suggests this effect is statistically significant. This is obviously not consistent with the nominal regression, as whose result shows non-monotone trend.

* housing Tower Block: The coefficient (-0.5010) for "housing Tower Block" suggests that living in a tower block, relative to the baseline, and among those with same level of contact, significantly increases the odds of being in a higher satisfaction category. This is consistent with the nominal regression.The positive coefficient and the t-value of 4.291 strongly indicate this is a significant predictor of higher satisfaction levels.

# 4 the largest discrepancies are between the observed frequencies and the expted frequencies.

```{r Pearson}
# residuals
pihat=predict(polr,df1,type='p')
m=rowSums(cbind(df2$low_satisfaction,df2$medium_satisfaction,df2$high_satisfaction))
res.pearson=(df2[,3:5]-pihat*m)/sqrt(pihat*m)
G=sum(res.pearson^2)

which(abs(res.pearson) == max(abs(res.pearson)), arr.ind = TRUE)

df2[4,]
```
The Pearson chisq-square statistic is `r G` and the high satisfaction with high contacts and housing in Tower Block has the largest discrepancies.  
